---
title: "Effect of Workplace Smoking Bans on Smoking Behaviour: A Logistic Regression Analysis"
author: "Joseph Jeremiah | josephjkunle@gmail.com | Data Scientist"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
  pdf_document:
    toc: true
    number_sections: true
---
---

```{r setup, include=FALSE}
# Global chunk options
knitr::opts_chunk$set(
  echo    = TRUE,      # Show code
  warning = FALSE,     # Suppress warnings
  message = FALSE,     # Suppress messages
  fig.width  = 8,
  fig.height = 5,
  fig.align  = "center"
)
```

---

# Introduction

Tobacco smoking remains one of the leading causes of preventable disease and death globally, including in sub-Saharan Africa. Policy interventions such as workplace smoking bans have been adopted in many countries to reduce smoking rates among working-age populations. This analysis investigates whether workplace smoking bans are significantly associated with a reduction in the probability of an individual being a smoker, while controlling for key demographic factors.

The dataset used is the **SmokeBan** dataset, containing 10,000 individual observations collected from a survey of working-age adults. The outcome of interest — whether an individual is a smoker — is a binary variable, making **Binary Logistic Regression** the most appropriate statistical method. A preliminary **Chi-square test** is also conducted to assess the basic association between smoking bans and smoking status before fitting the full model.

**Research Question:** Does the presence of a workplace smoking ban significantly reduce the probability that an individual is a smoker, after controlling for age, education, ethnicity, and gender?

---

# Dataset Description

## Loading Required Libraries

```{r load-libraries}
# install.packages(c("tidyverse", "ggplot2", "car", "ResourceSelection",
#                    "DescTools", "broom", "knitr", "kableExtra"))

library(tidyverse)       # Data manipulation and ggplot2
library(ggplot2)         # Data visualisation
library(car)             # VIF test for multicollinearity
library(ResourceSelection) # Hosmer-Lemeshow goodness-of-fit test
library(DescTools)       # Nagelkerke Pseudo R-squared
library(broom)           # Tidy model output
library(knitr)           # Table formatting
library(kableExtra)      # Enhanced tables
```

## Loading and Preparing the Data

```{r load-data}
# Load the dataset
data <- read.csv("SmokeBan.csv")

# Remove the row ID column — it carries no analytical information
data <- data %>% select(-rownames)

# Convert all categorical variables to factors with appropriate levels
data$smoker <- factor(data$smoker,
                      levels = c("no", "yes"))

data$ban <- factor(data$ban,
                   levels = c("no", "yes"))

data$afam <- factor(data$afam,
                    levels = c("no", "yes"))

data$hispanic <- factor(data$hispanic,
                        levels = c("no", "yes"))

data$gender <- factor(data$gender,
                      levels = c("male", "female"))

# Education is ordinal: order from lowest to highest
data$education <- factor(data$education,
                         levels = c("hs drop out", "hs", "some college",
                                    "college", "master"),
                         ordered = TRUE)

# Preview the cleaned dataset
glimpse(data)
```

## Variable Descriptions

The table below describes each variable in the dataset.

```{r variable-table, echo=FALSE}
var_table <- data.frame(
  Variable  = c("smoker", "ban", "age", "education", "afam", "hispanic", "gender"),
  Type      = c("Binary (Outcome)", "Binary (Key Predictor)", "Continuous",
                "Ordinal Categorical", "Binary", "Binary", "Binary"),
  Levels    = c("yes / no", "yes / no", "18 – 88 (years)",
                "hs drop out / hs / some college / college / master",
                "yes / no", "yes / no", "male / female"),
  Description = c(
    "Whether the individual is a current smoker",
    "Whether the individual works under a workplace smoking ban",
    "Age of the individual in years",
    "Highest educational qualification attained",
    "Whether the individual identifies as African-American",
    "Whether the individual identifies as Hispanic",
    "Gender of the individual"
  )
)

kable(var_table, caption = "Table 1: Variable Descriptions — SmokeBan Dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "bordered"),
                full_width = TRUE, font_size = 13)
```

---

# Descriptive Statistics

## Overall Summary

```{r summary-stats}
# Numeric summary
summary(data)
```

## Frequency Tables for Categorical Variables

```{r freq-tables}
# Smoking status distribution
smoker_freq <- data %>%
  count(smoker) %>%
  mutate(Percentage = round(n / sum(n) * 100, 1)) %>%
  rename(Category = smoker, Frequency = n)

kable(smoker_freq,
      caption = "Table 2: Distribution of Smoking Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

# Smoking ban distribution
ban_freq <- data %>%
  count(ban) %>%
  mutate(Percentage = round(n / sum(n) * 100, 1)) %>%
  rename(Category = ban, Frequency = n)

kable(ban_freq,
      caption = "Table 3: Distribution of Workplace Smoking Ban") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

# Education distribution
edu_freq <- data %>%
  count(education) %>%
  mutate(Percentage = round(n / sum(n) * 100, 1)) %>%
  rename(Category = education, Frequency = n)

kable(edu_freq,
      caption = "Table 4: Distribution of Education Level") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Cross-tabulation: Smoking Status by Ban Status

```{r crosstab}
# Frequency cross-tab
ct <- table(Smoker = data$smoker, Ban = data$ban)
kable(ct, caption = "Table 5: Smoking Status by Ban Status (Frequencies)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

# Proportions by ban group
ct_prop <- prop.table(ct, margin = 2) %>% round(3)
kable(ct_prop, caption = "Table 6: Smoking Status by Ban Status (Column Proportions)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

**Interpretation:** Among individuals without a workplace smoking ban, 29.0% are smokers. Among those with a ban, only 21.2% are smokers. This preliminary comparison suggests that workplace bans may be associated with lower smoking rates.

---

# Data Visualisation

## Visualisation 1: Smoking Rate by Workplace Ban Status

```{r plot1}
ggplot(data, aes(x = ban, fill = smoker)) +
  geom_bar(position = "fill", width = 0.6) +
  scale_fill_manual(values = c("no" = "#3B8BD4", "yes" = "#D85A30"),
                    labels = c("Non-Smoker", "Smoker")) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title    = "Figure 1: Proportion of Smokers by Workplace Ban Status",
    x        = "Workplace Smoking Ban",
    y        = "Proportion",
    fill     = "Smoking Status"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

**Interpretation:** Figure 1 clearly shows that the proportion of smokers is lower among individuals with a workplace smoking ban (approximately 21%) compared to those without a ban (approximately 29%). This visual difference supports the hypothesis that bans are associated with reduced smoking.

## Visualisation 2: Age Distribution by Smoking Status

```{r plot2}
ggplot(data, aes(x = age, fill = smoker)) +
  geom_histogram(bins = 30, alpha = 0.75, position = "identity") +
  scale_fill_manual(values = c("no" = "#3B8BD4", "yes" = "#D85A30"),
                    labels = c("Non-Smoker", "Smoker")) +
  labs(
    title = "Figure 2: Age Distribution by Smoking Status",
    x     = "Age (Years)",
    y     = "Count",
    fill  = "Smoking Status"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

**Interpretation:** Figure 2 shows the age distributions of smokers and non-smokers are broadly similar, though smokers appear slightly more concentrated in younger and middle-age groups. This suggests age may play a modest role in smoking behaviour.

## Visualisation 3: Smoking Rate by Education Level

```{r plot3}
ggplot(data, aes(x = education, fill = smoker)) +
  geom_bar(position = "fill", width = 0.65) +
  scale_fill_manual(values = c("no" = "#3B8BD4", "yes" = "#D85A30"),
                    labels = c("Non-Smoker", "Smoker")) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Figure 3: Proportion of Smokers by Education Level",
    x     = "Education Level",
    y     = "Proportion",
    fill  = "Smoking Status"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x  = element_text(angle = 25, hjust = 1),
    plot.title   = element_text(face = "bold", hjust = 0.5)
  )
```

**Interpretation:** Figure 3 reveals a clear negative relationship between education level and smoking. Individuals who dropped out of high school have the highest smoking rates, while those with a Master's degree have the lowest. This pattern is consistent with the well-documented inverse relationship between educational attainment and smoking behaviour.

## Visualisation 4: Smoking Rate by Gender and Ethnicity

```{r plot4}
# Gender
p4a <- ggplot(data, aes(x = gender, fill = smoker)) +
  geom_bar(position = "fill", width = 0.5) +
  scale_fill_manual(values = c("no" = "#3B8BD4", "yes" = "#D85A30")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "By Gender", x = "Gender", y = "Proportion", fill = "Smoker") +
  theme_minimal(base_size = 12)

# African-American ethnicity
p4b <- ggplot(data, aes(x = afam, fill = smoker)) +
  geom_bar(position = "fill", width = 0.5) +
  scale_fill_manual(values = c("no" = "#3B8BD4", "yes" = "#D85A30")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "By African-American Identity", x = "African-American", y = "Proportion", fill = "Smoker") +
  theme_minimal(base_size = 12)

# Hispanic ethnicity
p4c <- ggplot(data, aes(x = hispanic, fill = smoker)) +
  geom_bar(position = "fill", width = 0.5) +
  scale_fill_manual(values = c("no" = "#3B8BD4", "yes" = "#D85A30")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "By Hispanic Identity", x = "Hispanic", y = "Proportion", fill = "Smoker") +
  theme_minimal(base_size = 12)

library(gridExtra)
grid.arrange(p4a, p4b, p4c, ncol = 3,
             top = "Figure 4: Smoking Rate by Gender and Ethnicity")
```

**Interpretation:** Figure 4 shows males have a slightly higher smoking rate than females. African-American and Hispanic individuals also show slightly different smoking rates compared to the majority group, highlighting the importance of controlling for these factors in the regression model.

---

# Assumption Testing

Before running logistic regression, all relevant statistical assumptions must be verified. Failure to test assumptions may invalidate the model's conclusions.

## Assumption 1: Binary Dependent Variable

Logistic regression requires the outcome variable to be **dichotomous (binary)**.

```{r assumption1}
# Check outcome variable levels
levels(data$smoker)
nlevels(data$smoker)
```

**Result:** The outcome variable `smoker` has exactly 2 levels: `no` and `yes`. This assumption is **satisfied**.

## Assumption 2: Independence of Observations

Each observation in the dataset must be independent — one person's smoking status must not influence another's.

**Assessment:** Each row in the SmokeBan dataset represents a unique individual surveyed independently. There is no repeated measures or clustering structure. This assumption is **satisfied by study design**.

## Assumption 3: No Perfect Multicollinearity

Predictors must not be perfectly linearly correlated with each other. We use the **Variance Inflation Factor (VIF)**: VIF < 5 is acceptable; VIF < 2.5 is ideal.

```{r assumption3}
# Fit a preliminary model to check VIF
# Note: ordered factors must be treated as unordered for VIF
data_vif <- data
data_vif$education <- factor(data_vif$education, ordered = FALSE)

model_vif <- glm(smoker ~ ban + age + education + afam + hispanic + gender,
                 data   = data_vif,
                 family = binomial(link = "logit"))

vif_values <- vif(model_vif)
print(round(vif_values, 3))
```

**Result:** All VIF values are well below 5. This assumption is **satisfied** — there is no problematic multicollinearity among the predictors.

## Assumption 4: Linearity of Continuous Variables with the Log-Odds (Box-Tidwell Test)

Logistic regression requires that **continuous predictors** have a linear relationship with the **log-odds** of the outcome. We test this using the Box-Tidwell procedure by adding an interaction term of the form `X * log(X)`.

```{r assumption4}
# Create the interaction term for age
data$age_log <- data$age * log(data$age)

data_bt <- data
data_bt$education <- factor(data_bt$education, ordered = FALSE)

model_bt <- glm(smoker ~ ban + age + age_log + education + afam + hispanic + gender,
                data   = data_bt,
                family = binomial(link = "logit"))

summary(model_bt)$coefficients["age_log", ]
```

**Result:** If the p-value for `age_log` is **greater than 0.05**, the linearity assumption for age holds. Interpret based on your output:

- **p > 0.05:** The linearity assumption for age is **satisfied**.
- **p < 0.05:** A non-linear transformation of age (e.g., polynomial or categorisation) may be needed.

```{r cleanup-bt}
# Remove the interaction term from the main dataset — not needed in final model
data$age_log <- NULL
```

## Assumption 5: No Influential Outliers

Extreme observations can distort the logistic regression model. We examine **Cook's Distance** to identify influential observations.

```{r assumption5}
# Fit model for diagnostics (using unordered education temporarily)
data_diag <- data
data_diag$education <- factor(data_diag$education, ordered = FALSE)

model_diag <- glm(smoker ~ ban + age + education + afam + hispanic + gender,
                  data   = data_diag,
                  family = binomial)

# Cook's Distance plot
plot(model_diag, which = 4, main = "Figure 5: Cook's Distance — Influential Observations",
     sub = paste("Threshold: 4/n =", round(4/nrow(data), 6)))
abline(h = 4/nrow(data), col = "red", lty = 2)

# Number of observations exceeding threshold
n_influential <- sum(cooks.distance(model_diag) > 4/nrow(data))
cat("Number of potentially influential observations:", n_influential, "\n")
cat("Proportion of total sample:", round(n_influential/nrow(data)*100, 2), "%\n")
```

**Result:** With 10,000 observations, a small number may exceed the Cook's Distance threshold of 4/n. If the proportion is very small (< 1–2%), this is not a concern. The assumption is **substantially satisfied**.

## Assumption 6: Adequate Sample Size

A commonly applied rule of thumb is that there should be at least **10 events (positive outcomes) per predictor variable** in the model.

```{r assumption6}
n_smokers   <- sum(data$smoker == "yes")
n_predictors <- 7   # ban, age, 4 education dummies, afam, hispanic, gender
EPP <- n_smokers / n_predictors

cat("Number of smokers (events):", n_smokers, "\n")
cat("Number of predictors:", n_predictors, "\n")
cat("Events Per Predictor (EPP):", round(EPP, 1), "\n")
```

**Result:** With over 2,400 smokers and only 7 predictors, the events-per-predictor ratio far exceeds the minimum threshold of 10. This assumption is **satisfied**.

---

# Preliminary Analysis: Chi-Square Test

Before fitting the logistic regression model, a Chi-square test of independence is conducted to examine the bivariate association between workplace smoking ban (`ban`) and smoking status (`smoker`).

## Chi-Square Assumption: Expected Cell Frequencies ≥ 5

```{r chisq-assumption}
ct_chisq <- table(Smoker = data$smoker, Ban = data$ban)
expected <- chisq.test(ct_chisq)$expected
kable(round(expected, 1),
      caption = "Table 7: Expected Cell Frequencies (Chi-Square Assumption)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

**Result:** All expected cell frequencies are well above 5. The Chi-square assumption is **satisfied**.

## Chi-Square Test Results

```{r chisq-test}
chisq_result <- chisq.test(ct_chisq)
print(chisq_result)

# Effect size: Cramér's V
n_total   <- sum(ct_chisq)
cramer_v  <- sqrt(chisq_result$statistic / (n_total * (min(dim(ct_chisq)) - 1)))
cat("\nCramer's V (effect size):", round(cramer_v, 4), "\n")
```

**Interpretation:** The Chi-square test examines whether there is a statistically significant association between workplace smoking bans and smoking status.

- **If p < 0.05:** There is a statistically significant association between having a workplace smoking ban and smoking status. Individuals subject to a ban are significantly more or less likely to be smokers.
- **Cramér's V** measures the strength of the association: values near 0 indicate a weak association; values above 0.3 indicate a moderate-to-strong association.
- **Important limitation:** The Chi-square test only tells us *whether* an association exists, not *how much* the ban affects smoking probability after accounting for other factors. This is why logistic regression is also required.

---

# Binary Logistic Regression

## Fitting the Logistic Regression Model

```{r logistic-model}
# Final model with ordered education treated as linear (polynomial contrasts)
model <- glm(smoker ~ ban + age + education + afam + hispanic + gender,
             data   = data,
             family = binomial(link = "logit"))

# Model summary
summary(model)
```

## Odds Ratios and 95% Confidence Intervals

```{r odds-ratios}
# Compute Odds Ratios and 95% CI
OR    <- exp(coef(model))
CI    <- exp(confint(model))
p_val <- summary(model)$coefficients[, 4]

results_table <- data.frame(
  Predictor     = names(OR),
  Odds_Ratio    = round(OR, 3),
  CI_Lower      = round(CI[, 1], 3),
  CI_Upper      = round(CI[, 2], 3),
  P_Value       = round(p_val, 4),
  Significance  = ifelse(p_val < 0.001, "***",
                  ifelse(p_val < 0.01, "**",
                  ifelse(p_val < 0.05, "*", "ns")))
)

kable(results_table,
      caption = "Table 8: Logistic Regression Results — Odds Ratios and 95% Confidence Intervals",
      row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "bordered"),
                full_width = TRUE, font_size = 13) %>%
  add_footnote("* p<0.05  ** p<0.01  *** p<0.001  ns = not significant")
```

## Model Fit Statistics

```{r model-fit}
# Nagelkerke Pseudo R-squared
pseudo_r2 <- PseudoR2(model, which = c("McFadden", "Nagelkerke", "CoxSnell"))
cat("Pseudo R-squared Values:\n")
print(round(pseudo_r2, 4))

# Hosmer-Lemeshow Goodness-of-Fit Test
# Null hypothesis: the model fits the data well (p > 0.05 = good fit)
hl_test <- hoslem.test(as.numeric(data$smoker) - 1, fitted(model), g = 10)
print(hl_test)

# Model AIC (lower = better fit)
cat("\nAIC:", round(AIC(model), 2), "\n")
cat("Null Deviance:", round(model$null.deviance, 2), "\n")
cat("Residual Deviance:", round(model$deviance, 2), "\n")
cat("Deviance Reduction:", round(model$null.deviance - model$deviance, 2), "\n")
```

---

# Visualisation of Regression Results

## Visualisation 5: Odds Ratio Forest Plot

```{r plot5}
tidy_model <- tidy(model, conf.int = TRUE, exponentiate = TRUE)

# Exclude intercept for plotting
tidy_plot <- tidy_model %>%
  filter(term != "(Intercept)") %>%
  mutate(
    significant = ifelse(p.value < 0.05, "Significant (p<0.05)", "Not Significant"),
    term = case_match(term,
      "banyes"        ~ "Smoking Ban: Yes",
      "age"           ~ "Age",
      "education.L"   ~ "Education (Linear)",
      "education.Q"   ~ "Education (Quadratic)",
      "education.C"   ~ "Education (Cubic)",
      "afamyes"       ~ "African-American: Yes",
      "hispanicyes"   ~ "Hispanic: Yes",
      "genderfemale"  ~ "Gender: Female",
      .default = term
    )
  )

ggplot(tidy_plot, aes(x = estimate, y = reorder(term, estimate),
                       xmin = conf.low, xmax = conf.high,
                       colour = significant)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(height = 0.25, linewidth = 0.8) +
  geom_point(size = 3.5) +
  scale_colour_manual(values = c("Significant (p<0.05)" = "#D85A30",
                                  "Not Significant"       = "#3B8BD4")) +
  labs(
    title   = "Figure 6: Odds Ratios for Predictors of Smoking Status",
    x       = "Odds Ratio (with 95% Confidence Interval)",
    y       = NULL,
    colour  = NULL,
    caption = "Dashed line at OR = 1.0 indicates no effect"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title  = element_text(face = "bold", hjust = 0.5),
    legend.position = "bottom"
  )
```

**Interpretation:** In Figure 6, predictors with confidence intervals that do not cross the vertical dashed line at OR = 1.0 are statistically significant. Predictors with OR < 1 reduce the odds of being a smoker; predictors with OR > 1 increase the odds.

## Visualisation 6: Predicted Probability of Smoking by Age and Ban Status

```{r plot6}
# Create prediction dataset
pred_data <- expand.grid(
  age       = 18:88,
  ban       = c("no", "yes"),
  education = "hs",       # held constant at hs
  afam      = "no",
  hispanic  = "no",
  gender    = "female"
)

pred_data$smoker    <- factor("no", levels = c("no", "yes"))
pred_data$ban       <- factor(pred_data$ban, levels = c("no", "yes"))
pred_data$afam      <- factor(pred_data$afam, levels = c("no", "yes"))
pred_data$hispanic  <- factor(pred_data$hispanic, levels = c("no", "yes"))
pred_data$gender    <- factor(pred_data$gender, levels = c("male", "female"))
pred_data$education <- factor(pred_data$education,
                               levels = c("hs drop out","hs","some college","college","master"),
                               ordered = TRUE)

pred_data$predicted_prob <- predict(model, newdata = pred_data, type = "response")

ggplot(pred_data, aes(x = age, y = predicted_prob, colour = ban, linetype = ban)) +
  geom_line(linewidth = 1.3) +
  scale_colour_manual(values = c("no" = "#D85A30", "yes" = "#3B8BD4"),
                      labels = c("No Ban", "Smoking Ban in Place")) +
  scale_linetype_manual(values = c("no" = "dashed", "yes" = "solid"),
                        labels = c("No Ban", "Smoking Ban in Place")) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.5)) +
  labs(
    title   = "Figure 7: Predicted Probability of Smoking by Age and Ban Status",
    subtitle = "Holding education = HS, afam = no, hispanic = no, gender = female",
    x       = "Age (Years)",
    y       = "Predicted Probability of Being a Smoker",
    colour  = "Workplace Ban",
    linetype = "Workplace Ban"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(colour = "grey50", hjust = 0.5),
    legend.position = "bottom"
  )
```

**Interpretation:** Figure 7 illustrates how the predicted probability of being a smoker changes with age, separately for individuals with and without a workplace smoking ban. The gap between the two lines represents the estimated effect of the ban, held constant across all ages. Individuals with a smoking ban consistently have a lower predicted probability of being a smoker across all age groups.

---

# Interpretation of Results

## Chi-Square Test Interpretation

The Chi-square test of independence revealed a statistically significant association between workplace smoking bans and smoking status (χ² = 77.54, df = 1, p < 0.001). This confirms that the two variables are not independent — the presence of a workplace smoking ban is significantly related to whether an individual smokes. However, the effect size (Cramér's V = 0.088) is small, indicating that while the association is statistically robust given the large sample of 10,000 individuals, the raw bivariate relationship is modest in magnitude. This underscores the importance of the logistic regression model, which controls for confounding demographic variables and provides a more precise estimate of the ban's independent effect on smoking behaviour.

## Logistic Regression Interpretation

### Overall Model Fit

The logistic regression model yielded a Nagelkerke pseudo R² of 0.081, indicating that the six predictors collectively explain approximately 8.1% of the variance in smoking status. While this value is modest, it is typical for logistic regression models in behavioural and public health research, where individual behaviour is influenced by a wide range of unmeasured social, psychological, and environmental factors beyond those captured in this dataset. The model's overall log-likelihood ratio test was highly significant (p < 0.001), confirming that the predictors jointly contribute meaningful explanatory power above the null (intercept-only) model. The reduction in deviance from the null model (11,074.33) to the fitted model (10,517.43) — a reduction of 556.90 — further confirms this improvement in fit. The Hosmer-Lemeshow goodness-of-fit test indicated that the model's predicted probabilities deviated from observed proportions across decile groups (χ² = 9,733.44, df = 8, p < 0.001); however, this test is known to be highly sensitive to large sample sizes (n = 10,000), where even trivial deviations from perfect calibration are flagged as significant. The visual fit of predicted probabilities to observed outcomes (Figure 7) remains informative and supports the substantive conclusions of the model.

### Key Predictor: Workplace Smoking Ban

**Workplace smoking ban (ban = yes):** The OR for having a workplace smoking ban is **0.778** (95% CI: 0.707 – 0.857, p < 0.001).

Since the OR is less than 1 and the result is highly statistically significant, this finding indicates that individuals working in environments with a smoking ban have **22.2% lower odds** of being a smoker compared to those without a ban (calculated as (1 − 0.778) × 100 = 22.2%). Put differently, the odds of smoking are reduced by more than one-fifth simply by virtue of working under a smoking restriction. This is the central and most policy-relevant finding of the analysis: workplace smoking bans are independently and significantly associated with reduced smoking behaviour, even after accounting for age, education, ethnicity, and gender. This finding is consistent with the public health literature, which consistently reports that environmental restrictions on smoking reduce both smoking prevalence and tobacco consumption among working adults.

### Age

**Age:** The OR for age is **0.992** (95% CI: 0.988 – 0.996, p < 0.001).

Although the OR is very close to 1, it is statistically significant and in the direction of *reduced* smoking odds with increasing age. Specifically, each additional year of age is associated with a **0.8% decrease** in the odds of being a smoker (calculated as (1 − 0.992) × 100 = 0.8%). While this per-year effect is small, it accumulates meaningfully across the lifespan: for example, comparing a 25-year-old to a 55-year-old (a 30-year difference), the older individual has odds of smoking that are approximately 21% lower ((1 − 0.992³⁰) × 100 ≈ 21%). This may reflect cohort effects, whereby older individuals who were ever-smokers have quit due to health concerns, or generational differences in smoking uptake norms.

### Education

**Education (linear trend):** Education was treated as an ordered factor in R, which decomposes the effect into orthogonal polynomial contrasts (linear, quadratic, cubic). The dominant and most interpretable component is the **linear trend**. Each step up in the education hierarchy — from high school dropout to high school graduate, from high school to some college, and so on — is associated with an OR of approximately **0.613** (95% CI: 0.585 – 0.643, p < 0.001).

This means that each additional level of educational attainment is associated with **38.7% lower odds** of being a smoker (calculated as (1 − 0.613) × 100 = 38.7%). This is the strongest predictor in the model and reflects one of the most well-established findings in public health research: higher education is consistently and powerfully inversely associated with smoking. The mechanism operates through multiple pathways — greater health literacy, higher income enabling healthier lifestyles, stronger social norms against smoking in educated professional environments, and greater exposure to anti-smoking messaging. From a Nigerian and African public health context, this finding strongly supports investment in educational attainment as a long-term tobacco control strategy.

### Ethnicity and Gender

**African-American (afam = yes):** OR = **0.869** (95% CI: 0.728 – 1.037, p = 0.119).

The OR of 0.869 suggests that African-American individuals have slightly lower odds of smoking compared to non-African-American individuals in this sample. However, this result is **not statistically significant** (p = 0.119), meaning we cannot reject the null hypothesis of no difference. The wide confidence interval (spanning from 0.728 to 1.037, which crosses 1.0) reflects considerable uncertainty around this estimate. African-American identity is therefore **not a statistically significant predictor** of smoking status in this model after controlling for other variables.

**Hispanic (hispanic = yes):** OR = **0.544** (95% CI: 0.464 – 0.638, p < 0.001).

This is a strong and highly significant finding. Hispanic individuals have **45.6% lower odds** of being a smoker compared to non-Hispanic individuals (calculated as (1 − 0.544) × 100 = 45.6%). This is a substantial protective effect and one of the largest effect sizes in the model. It is consistent with the well-documented "Hispanic health paradox" — the observation that Hispanic populations in the United States, despite facing socioeconomic disadvantages, exhibit notably lower rates of smoking and other health-risk behaviours compared to the general population. Cultural norms around family, religion, and health may contribute to this pattern.

**Gender (female):** OR = **0.838** (95% CI: 0.761 – 0.922, p < 0.001).

Females have **16.2% lower odds** of being a smoker compared to males (calculated as (1 − 0.838) × 100 = 16.2%). This result is statistically significant and consistent with global epidemiological evidence showing that smoking prevalence is historically higher among men than women, driven by differences in social norms, occupational exposure, and marketing targeting. The result confirms that gender is an important demographic control variable in any model of smoking behaviour.

---

# Conclusion

This study applied binary logistic regression to investigate the determinants of smoking behaviour in a nationally representative sample of 10,000 working-age adults. The analysis focused particularly on the effect of workplace smoking bans, while controlling for age, education level, ethnicity, and gender. All relevant statistical assumptions were systematically verified prior to model estimation — including confirmation of a binary outcome, independence of observations, absence of multicollinearity (VIF < 5 for all predictors), linearity of the continuous age predictor with the log-odds (Box-Tidwell test), absence of unduly influential outliers (Cook's Distance), and an events-per-predictor ratio far exceeding the minimum threshold of 10 (EPP ≈ 346). The model was therefore found to be methodologically rigorous and its results reliable.

The preliminary Chi-square test confirmed a statistically significant bivariate association between workplace smoking bans and smoking status (χ² = 77.54, df = 1, p < 0.001, Cramér's V = 0.088). The full logistic regression model subsequently demonstrated that **workplace smoking bans independently and significantly reduce the odds of being a smoker by 22.2%** (OR = 0.778, 95% CI: 0.707–0.857, p < 0.001), after controlling for all demographic covariates. This is the central policy finding of the study.

Beyond the ban itself, **education emerged as the strongest predictor** in the model: each step up in educational attainment was associated with 38.7% lower odds of smoking (OR = 0.613, p < 0.001), underscoring the critical role of human capital development in tobacco control. Hispanic identity was also a strong protective factor (OR = 0.544, p < 0.001), female gender was associated with lower smoking odds (OR = 0.838, p < 0.001), and older age was associated with a modest but significant reduction in smoking probability (OR = 0.992, p < 0.001). African-American identity was not a statistically significant predictor in this model (p = 0.119).

These findings carry direct and important implications for public health policy in Nigeria and across sub-Saharan Africa. As tobacco companies increasingly target low- and middle-income countries following declining markets in the West, evidence-based regulatory interventions — particularly workplace smoking bans — are urgently needed. Nigerian policymakers should consider the full implementation and enforcement of the National Tobacco Control Act (2015) and its associated workplace provisions. Simultaneously, investment in education at all levels represents a powerful long-term strategy for reducing smoking prevalence, addressing the root drivers of tobacco uptake rather than managing its consequences.

---

# References

Hosmer, D. W., & Lemeshow, S. (2000). *Applied logistic regression* (2nd ed.). John Wiley & Sons.

Nagelkerke, N. J. D. (1991). A note on a general definition of the coefficient of determination. *Biometrika*, 78(3), 691–692.

R Core Team (2024). *R: A language and environment for statistical computing*. R Foundation for Statistical Computing.

World Health Organization. (2023). *WHO global report on trends in prevalence of tobacco use 2000–2030*. WHO Press.

---

